---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# raretrans

[![Build Status](https://travis-ci.com/atyre2/raretrans.svg?token=fZ6v1nEUhWRrRW3uj9TH&branch=master)](https://travis-ci.com/atyre2/raretrans)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/atyre2/raretrans?branch=master&svg=true)](https://ci.appveyor.com/project/atyre2/raretrans)
[![codecov](https://codecov.io/gh/atyre2/raretrans/branch/master/graph/badge.svg?token=NWFl8m4owW)](https://codecov.io/gh/atyre2/raretrans)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/raretrans)](https://cran.r-project.org/package=raretrans)

Functions to create matrix population models from a combination
    of data on stage/age transitions and Bayesian prior information. This 
    compensates for structural problems caused by missing observatons of rare
    transitions. 

## Installation

raretrans is not currently available from CRAN, but you can install from github.

Code and data used to produce Raymond L. Tremblay, Andrew J. Tyre , Maria-Eglée Pérez, James D. Ackerman (in review) Population projections from holey matrices: Using prior information to estimate rare transition events is tagged v1.0.0, so:

```{r, eval = FALSE}
# install.packages("devtools") # if needed
devtools::install_github("atyre2/raretrans@v1.0.0")
```

Alternatively, [download a package archive file from the release page](https://github.com/atyre2/raretrans/releases). 

Get the development version from github with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("atyre2/raretrans")
```
If you want the vignettes, you have to add some additional arguments.

```{r, eval = FALSE}
devtools::install_github("atyre2/raretrans", build = TRUE, build_opts = c("--no-resave-data", "--no-manual"))
```

## Overview

There are two main functions in this package, `fill_transitions()` and `fill_fertility()`.
Both of them take a list of two matrices as their first argument, the population 
projection matrix decomposed into a matrix of the probability of transitions
between stages, and a matrix describing the contribution to the next census via
reproduction. These matrices can be constructed by hand, or created with 
`popbio::projection_matrix(..., TF = TRUE)` from a data frame with transitions
between stages for individuals. By default, they return a matrix 'filled in' with 
uninformative or uniform priors. Informative priors can be specified as well.

For detailed examples of using these functions with a single population see the
vignette [Working with a single population and transition](https://atyre2.github.io/raretrans/articles/onepopperiod.html).
The vignette [Effect of prior information on transition probabilities and fertilities](https://atyre2.github.io/raretrans/articles/transition_priors.html) 
shows how to work with a dataset of many populations and transitions, and 
produces all the figures for Tremblay et al. (submitted). 

For an overview of population projection models generally, see 
[Hal Caswell, 2001, Matrix Population Models: Construction, Analysis, and Interpretation](https://books.google.com/books/about/Matrix_Population_Models.html?id=CPsTAQAAIAAJ).


## Code of Conduct

Please note that the 'raretrans' project is released with a
[Contributor Code of Conduct](CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.
