---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# raretrans

[![Build Status](https://travis-ci.com/atyre2/raretrans.svg?token=fZ6v1nEUhWRrRW3uj9TH&branch=master)](https://travis-ci.com/atyre2/raretrans)
[![Coverage status](https://codecov.io/gh/atyre2/raretrans/branch/master/graph/badge.svg)](https://codecov.io/github/atyre2/raretrans?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/raretrans)](https://cran.r-project.org/package=raretrans)

Functions to create matrix population models from a combination
    of data on stage/age transitions and Bayesian prior information. This 
    compensates for structural problems caused by missing observatons of rare
    transitions. 

## Installation

raretrans is not currently available from CRAN, but you can install the development version from github with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("atyre2/raretrans")
```

