---
title: "Matrices of animal populations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Matrices of animal populations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Although the paper focuses on a dataset of a perennial plant, the same principles 
work for animal matrices too. Small sample sizes can comprimise the estimation of 
survival and reproduction, especially if the analyst believes that matrix entries
should vary by age, sex, or other environmental conditions. In some cases there
are no available data, and the analysis proceeds using "guesses", AKA prior beliefs.


```{r setup}
library(tidyverse)
library(popbio)
library(raretrans)

# Raymond's theme modifications
rlt_theme <- theme(axis.title.y = element_text(colour="grey20",size=15,face="bold"),
        axis.text.x = element_text(colour="grey20",size=15, face="bold"),
        axis.text.y = element_text(colour="grey20",size=15,face="bold"),  
        axis.title.x = element_text(colour="grey20",size=15,face="bold"))
```

# Snail kites

