---
title: "Matrices of animal populations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Matrices of animal populations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Although the paper focuses on a dataset of a perennial plant, the same principles 
work for animal matrices too. Small sample sizes can comprimise the estimation of 
survival and reproduction, especially if the analyst believes that matrix entries
should vary by age, sex, or other environmental conditions. In some cases there
are no available data, and the analysis proceeds using "guesses", AKA prior beliefs.
Using simple prior distributions for these parameters enables the calculation
of credible intervals for both matrix entries and derived quantities in a
natural way.


```{r setup}
library(tidyverse)
library(popbio)
library(raretrans)
library(huxtable)

# Raymond's theme modifications
rlt_theme <- theme(axis.title.y = element_text(colour="grey20",size=15,face="bold"),
        axis.text.x = element_text(colour="grey20",size=15, face="bold"),
        axis.text.y = element_text(colour="grey20",size=15,face="bold"),  
        axis.title.x = element_text(colour="grey20",size=15,face="bold"))
```

# Snail kites

Stephen Beissinger [-@beissinger1995snailkite] developed a PVA for the 
endangered snail kite in the Florida Everglades. A key piece of that 
problem was a periodic variation in water levels caused by management. Kites had
lower survival and reproduction when water levels were drawn down in 
dry years to support irrigation. Data on radio telemetered animals and nest
success were available for "high water" years, and anecdotal evidence was used
to guesstimate values for "drought" and "lag" years. The model used three age 
classes because "[a] finer age-structured model was not feasible due to small sample sizes of age-specific demographic data". Standard deviations and ranges were
provided for each of the parameters, although no information was given about 
how stochastic values were simulated. 
```{r}
b1995_table2 <- tribble(
  ~env_state, ~age_class, ~mean_ns, ~sd_ns, ~range_ns, ~prop_nesting, ~attempts, ~mean_s, ~sd_s, ~range_s,
  "", "", "Nesting success", "","", "", "", "Annual survival", "","",
  "Environmental state", "Age Class", "Mean", "SD","Range", "Proportion Nesting", "Nesting attempts per year", "Mean", "SD", "Range",
  "Drought year", "Fledgling", "0.00", "0.00", "0.00-0.00", "0.00", "0.00", "0.50","0.10", "0.30-0.80",
  "", "Subadult", "0.03", "0.10", "0.00-0.25", "0.15", "1.00", "0.60", "0.10", "0.40-0.90",
  "", "Adult", "0.03", "0.10", "0.00-0.25", "0.15", "1.00", "0.60", "0.10", "0.40-0.90",
  "Lag year", "Fledgling", "0.00", "0.00", "0.00-0.00", "0.00", "0.00", "0.85","0.03", "0.75-0.92",
  "", "Subadult", "0.16", "0.10", "0.04-0.33", "0.15", "1.00", "0.90", "0.03", "0.80-0.95",
  "", "Adult",  "0.16", "0.10", "0.04-0.33", "0.80", "1.50", "0.90", "0.03", "0.80-0.95",
  "High year", "Fledgling", "0.00", "0.00", "0.00-0.00", "0.00", "0.00", "0.90","0.03", "0.75-0.92",
  "", "Subadult", "0.30", "0.10", "0.1-0.40", "0.25", "1.00", "0.95", "0.03", "0.85-0.98",
  "", "Adult", "0.30", "0.10", "0.1-0.40", "1.00", "2.20", "0.95", "0.03", "0.85-0.98"
)
huxtable(b1995_table2) %>% 
  set_colspan(row = 1, col = c(3,8), value = 3) %>% 
  theme_article(header_col = FALSE) %>% 
  set_bottom_border(row = 1, col = everywhere, value = 0) %>% 
  set_bottom_border(row = 2, col = everywhere, value = 1)
```



## Survival values

In stage structured matrices generally, the survival component of the matrix is
a multinomial because of the possibility of transitioning between more than 2 
stages. In an age structured matrix like this one, the only possibilities are
survival to the next stage or death, so a simple $Beta(\alpha, \beta)$ prior
is all that is needed. The mean is given by $\alpha / \left(\alpha + \beta\right)$,
while the variance of the distribution is 

$$
var(X) = \frac{\alpha \beta}{\left(\alpha + \beta\right)^2\left(\alpha + \beta + 1\right)}
$$

We have a mean and standard deviation
