---
title: "Matrices of animal populations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Matrices of animal populations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Although the paper focuses on a dataset of a perennial plant, the same principles 
work for animal matrices too. Small sample sizes can comprimise the estimation of 
survival and reproduction, especially if the analyst believes that matrix entries
should vary by age, sex, or other environmental conditions. In some cases there
are no available data, and the analysis proceeds using "guesses", AKA prior beliefs.
Using simple prior distributions for these parameters enables the calculation
of credible intervals for both matrix entries and derived quantities in a
natural way.


```{r setup}
library(tidyverse)
library(popbio)
library(raretrans)

# Raymond's theme modifications
rlt_theme <- theme(axis.title.y = element_text(colour="grey20",size=15,face="bold"),
        axis.text.x = element_text(colour="grey20",size=15, face="bold"),
        axis.text.y = element_text(colour="grey20",size=15,face="bold"),  
        axis.title.x = element_text(colour="grey20",size=15,face="bold"))
```

# Snail kites

Stephen Beissinger [-@beissinger1995snailkite] developed a PVA for the 
endangered snail kite in the Florida Everglades. A key piece of that 
problem was a periodic variation in water levels caused by management. Kites had
lower survival and reproduction when water levels were drawn down in 
dry years to support irrigation.
