---
title: "Matrices of animal populations"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Matrices of animal populations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

Although the paper focuses on a dataset of a perennial plant, the same principles 
work for animal matrices too. Small sample sizes can comprimise the estimation of 
survival and reproduction, especially if the analyst believes that matrix entries
should vary by age, sex, or other environmental conditions. In some cases there
are no available data, and the analysis proceeds using "guesses", AKA prior beliefs.
Using simple prior distributions for these parameters enables the calculation
of credible intervals for both matrix entries and derived quantities in a
natural way.


```{r setup}
library(tidyverse)
library(popbio)
library(raretrans)
library(huxtable)

# Raymond's theme modifications
rlt_theme <- theme(axis.title.y = element_text(colour="grey20",size=15,face="bold"),
        axis.text.x = element_text(colour="grey20",size=15, face="bold"),
        axis.text.y = element_text(colour="grey20",size=15,face="bold"),  
        axis.title.x = element_text(colour="grey20",size=15,face="bold"))
```

# Snail kites

Stephen Beissinger [-@beissinger1995snailkite] developed a PVA for the 
endangered snail kite in the Florida Everglades. A key piece of that 
problem was a periodic variation in water levels caused by management. Kites had
lower survival and reproduction when water levels were drawn down in 
dry years to support irrigation. Data on radio telemetered animals and nest
success were available for "high water" years, and anecdotal evidence was used
to guesstimate values for "drought" and "lag" years. The model used three age 
classes because "[a] finer age-structured model was not feasible due to small sample sizes of age-specific demographic data". Standard deviations and ranges were
provided for each of the parameters, although no information was given about 
how stochastic values were simulated. 
```{r}
b1995_table2 <- tribble(
  ~env_state, ~age_class, ~mean_ns, ~sd_ns, ~range_ns, ~prop_nesting, ~attempts, ~mean_s, ~sd_s, ~range_s,
  "", "", "Nesting success", "","", "", "", "Annual survival", "","",
  "Environmental state", "Age Class", "Mean", "SD","Range", "Proportion Nesting", "Nesting attempts per year", "Mean", "SD", "Range",
  "Drought year", "Fledgling", "0.00", "0.00", "0.00-0.00", "0.00", "0.00", "0.50","0.10", "0.30-0.80",
  "", "Subadult", "0.03", "0.10", "0.00-0.25", "0.15", "1.00", "0.60", "0.10", "0.40-0.90",
  "", "Adult", "0.03", "0.10", "0.00-0.25", "0.15", "1.00", "0.60", "0.10", "0.40-0.90",
  "Lag year", "Fledgling", "0.00", "0.00", "0.00-0.00", "0.00", "0.00", "0.85","0.03", "0.75-0.92",
  "", "Subadult", "0.16", "0.10", "0.04-0.33", "0.15", "1.00", "0.90", "0.03", "0.80-0.95",
  "", "Adult",  "0.16", "0.10", "0.04-0.33", "0.80", "1.50", "0.90", "0.03", "0.80-0.95",
  "High year", "Fledgling", "0.00", "0.00", "0.00-0.00", "0.00", "0.00", "0.90","0.03", "0.75-0.92",
  "", "Subadult", "0.30", "0.10", "0.1-0.40", "0.25", "1.00", "0.95", "0.03", "0.85-0.98",
  "", "Adult", "0.30", "0.10", "0.1-0.40", "1.00", "2.20", "0.95", "0.03", "0.85-0.98"
)
huxtable(b1995_table2) %>% 
  set_colspan(row = 1, col = c(3,8), value = 3) %>% 
  theme_article(header_col = FALSE) %>% 
  set_bottom_border(row = 1, col = everywhere, value = 0) %>% 
  set_bottom_border(row = 2, col = everywhere, value = 1) %>% 
  set_caption("Table 1: Snail kite vital rates from [@beissinger1995snailkite], Table 2.")
```



## Survival values

In stage structured matrices generally, the survival component of the matrix is
a multinomial because of the possibility of transitioning between more than 2 
stages. In an age structured matrix like this one, the only possibilities are
survival to the next stage or death, so a simple $Beta(\alpha, \beta)$ prior
is all that is needed. The mean is given by $\alpha / \left(\alpha + \beta\right)$,
while the variance of the distribution is 

$$
var(X) = \frac{\alpha \beta}{\left(\alpha + \beta\right)^2\left(\alpha + \beta + 1\right)}
$$

We have a mean and standard deviation, and can obtain approximate $\alpha$ and $\beta$ as follows

$$
\begin{align*}
\nu = & \frac{\mu\left(1-\mu\right)}{var} \\
\alpha = & \nu\mu \\
\beta = & \left(1-\mu\right) \nu
\end{align*}

$$

as long as $\mu\left(1-\mu\right) < var$. This condition is met for all the values in 
Beissinger's Table 2. 

```{r}
kite_parms <- b1995_table2 %>% 
  slice(-(1:2)) %>% # remove top two rows
  separate(col = range_ns, into = c("min_ns","max_ns"), sep = "-") %>% 
  separate(col = range_s, into = c("min_s","max_s"), sep = "-") %>% 
  mutate_at(.vars = 3:12, .funs = as.numeric) %>% # convert to numeric
  mutate(nu_ns = mean_ns * (1 - mean_ns) / sd_ns^2,
         alpha_ns = mean_ns * nu_ns,
         beta_ns = (1-mean_ns)*nu_ns,
         nu_s = mean_s * (1 - mean_s) / sd_s^2,
         alpha_s = mean_s * nu_s,
         beta_s = (1-mean_s)*nu_s) %>% 
  mutate_if(is.numeric, ~ifelse(is.nan(.x), NA_real_, .x))
kite_parms %>% 
  select(env_state, age_class, alpha_ns, beta_ns, alpha_s, beta_s) %>% 
  mutate_if(is.numeric, ~format(.x, digits = 2)) %>% 
  bind_rows(tibble(env_state = c("","Environmental State"),
                   age_class = c("", "Age Class"),
                   alpha_ns = c("Nest Success", "$\\alpha$"),
                   beta_ns = c("", "$\\beta$"),
                   alpha_s = c("Survival", "$\\alpha$"),
                   beta_s = c("", "$\\beta$")), .) %>% 
  hux() %>% 
  set_escape_contents(row = 2, col = 3:6, value = FALSE) %>% 
  theme_article(header_col = FALSE) %>% 
  set_na_string(value = "") %>% 
  set_caption("Table 2: Values of $\\alpha$ and $\\beta$ for each age class and environmental state.")
  
```

One of the benefits of articulating the beliefs this way is that the reader can
immediately see how much confidence the analyst has placed in each of the parameter estimates.
The sum of $\alpha$ and $\beta$ is effective sample size. So we have less confidence in the 
estimates of nest success in lag and drought years than high water years, which is consistent
what is known. However, within each environmental state we have more confidence in the survival
estimates of fledgling than subadult or adult stages, and less confidence in
estimates from high water years than lag years. Both of these are inconsistent with
prior beliefs, but are a consequence of fixing the standard deviation of the estimates. 
Fixing the standard deviation gives a consistent belief about
precision if the parameter is normally distributed, because the mean and standard deviation
are independent. This is not true of a Beta distribution. 

```{r}
kite_parms2 <- kite_parms %>% 
  mutate(env_state = rep(env_state[c(1,4,7)], each = 3),
         n_ns = alpha_ns + beta_ns,
         n_s = alpha_s + beta_s,
         lwr_ns = pbeta(min_ns, alpha_ns, beta_ns),
         upr_ns = pbeta(max_ns, alpha_ns, beta_ns),
         lwr_s = pbeta(min_s, alpha_s, beta_s),
         upr_s = pbeta(max_s, alpha_s, beta_s),
         plot_ns = map2(alpha_ns, beta_ns, ~tibble(x = seq(0,1,0.01),
                                                   d_ns = dbeta(x, .x, .y))),
         plot_s = map2(alpha_s, beta_s, ~tibble(x = seq(0,1,0.01),
                                                d_s = dbeta(x, .x, .y)))) %>% 
  unnest() %>% 
  mutate(env_state = factor(env_state, levels = c("Drought year","Lag year", "High year")),
         age_class = factor(age_class, levels = c("Fledgling", "Subadult", "Adult")))
ggplot(data = filter(kite_parms2, age_class != "Fledgling"),
       mapping = aes(x = x, y = d_ns)) + 
  geom_area() + 
  facet_grid(env_state~age_class) +
  labs(x = "Nest success", y = "Density")+
  theme_classic()
```

```{r}
ggplot(data = filter(kite_parms2),
       mapping = aes(x = x, y = d_s)) + 
  geom_area() + 
  facet_grid(env_state~age_class) +
  labs(x = "Nest success", y = "Density")+
  theme_classic()
```

# Literature cited
