---
title: "Matrices of animal populations"
output: 
  rmarkdown::pdf_document:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Matrices of animal populations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
options(huxtable.add_colnames = FALSE,
        digits = 3)
```

Although the paper focuses on a perennial plant data set, the same principles 
work for animal matrices too. Small sample sizes can compromise the estimation of 
survival and reproduction, especially if the analyst believes that matrix entries
should vary by age, sex, or other environmental conditions. In some cases there
are no available data, and the analysis proceeds using "guesses", AKA prior beliefs.
Using simple prior distributions for these parameters enables the calculation
of credible intervals for both matrix entries and derived quantities in a
natural way.

This example does not use the `raretrans` functions, because the information in
Beissinger [-@beissinger1995snailkite] is provided as means and standard 
deviations, not as numbers of transitions. 


```{r setup}
library(tidyverse)
library(popbio)
library(huxtable)

# Raymond's theme modifications
rlt_theme <- theme(axis.title.y = element_text(colour="grey20",size=15,face="bold"),
        axis.text.x = element_text(colour="grey20",size=15, face="bold"),
        axis.text.y = element_text(colour="grey20",size=15,face="bold"),  
        axis.title.x = element_text(colour="grey20",size=15,face="bold"))
```

# Snail kites

Stephen Beissinger [-@beissinger1995snailkite] developed a PVA for the 
endangered snail kite in the Florida Everglades. A key piece of that 
problem was a periodic variation in water levels caused by management. Kites had
lower survival and reproduction when water levels were drawn down in 
dry years to support irrigation. Data on radio telemetered animals and nest
success were available for "high water" years, and anecdotal evidence was used
to guesstimate values for "drought" and "lag" years. The model used three age 
classes because "[a] finer age-structured model was not feasible due to small sample sizes of age-specific demographic data". Standard deviations and ranges were
provided for some of the parameters. The methods describe survival values as drawn from
a normal distribution. 

```{r}
b1995_table2 <- tribble(
  ~env_state, ~age_class, ~mean_ns, ~sd_ns, ~range_ns, ~prop_nesting, ~attempts, ~mean_s, ~sd_s, ~range_s,
  "", "", "Nesting success", "","", "", "", "Annual survival", "","",
  "Environmental state", "Age Class", "Mean", "SD","Range", "Proportion Nesting", 
  "Nesting attempts per year", "Mean", "SD", "Range",
  "Drought year", "Fledgling", "0.00", "0.00", "0.00-0.00", "0.00", "0.00", "0.50","0.10", "0.30-0.80",
  "", "Subadult", "0.03", "0.10", "0.00-0.25", "0.15", "1.00", "0.60", "0.10", "0.40-0.90",
  "", "Adult", "0.03", "0.10", "0.00-0.25", "0.15", "1.00", "0.60", "0.10", "0.40-0.90",
  "Lag year", "Fledgling", "0.00", "0.00", "0.00-0.00", "0.00", "0.00", "0.85","0.03", "0.75-0.92",
  "", "Subadult", "0.16", "0.10", "0.04-0.33", "0.15", "1.00", "0.90", "0.03", "0.80-0.95",
  "", "Adult",  "0.16", "0.10", "0.04-0.33", "0.80", "1.50", "0.90", "0.03", "0.80-0.95",
  "High year", "Fledgling", "0.00", "0.00", "0.00-0.00", "0.00", "0.00", "0.90","0.03", "0.75-0.92",
  "", "Subadult", "0.30", "0.10", "0.1-0.40", "0.25", "1.00", "0.95", "0.03", "0.85-0.98",
  "", "Adult", "0.30", "0.10", "0.1-0.40", "1.00", "2.20", "0.95", "0.03", "0.85-0.98")

huxtable(b1995_table2) %>% 
  set_colspan(row = 1, col = c(3,8), value = 3) %>% 
  theme_article(header_col = FALSE) %>% 
  set_width(0.9) %>% 
  set_col_width(c(0.15, 0.1, 0.075, 0.075, 0.1, 0.15, 0.15, 0.075, 0.075, 0.1)) %>% 
  set_position("left") %>% 
  set_bottom_border(row = 1, col = everywhere, value = 0) %>% 
  set_bottom_border(row = 2, col = everywhere, value = 1) %>% 
  set_caption("Snail kite vital rates from Beissinger (1995), Table 2.")
```

## Survival values

In stage structured matrices generally, the survival component of the matrix is
a multinomial because of the possibility of transitioning between more than 2 
stages. In an age structured matrix like this one, the only possibilities are
survival to the next stage or death, so a simple $Beta(\alpha, \beta)$ prior
is all that is needed. The mean is given by $\alpha / \left(\alpha + \beta\right)$,
while the variance of the distribution is 

$$
\sigma^2 = \frac{\alpha \beta}{\left(\alpha + \beta\right)^2\left(\alpha + \beta + 1\right)}
$$

We have a mean and standard deviation from Table 1, and can obtain approximate $\alpha$ and $\beta$ as follows


\begin{align*}
\nu = & \frac{\mu\left(1-\mu\right)}{\sigma^2} \\
\alpha = & \nu\mu \\
\beta = & \left(1-\mu\right) \nu
\end{align*}



as long as $\mu\left(1-\mu\right) < \sigma^2$. This condition is met for all the values in 
Beissinger's Table 2. 

```{r}
kite_parms <- b1995_table2 %>% 
  slice(-(1:2)) %>% # remove top two rows
  separate(col = range_ns, into = c("min_ns","max_ns"), sep = "-") %>% 
  separate(col = range_s, into = c("min_s","max_s"), sep = "-") %>% 
  mutate(across(.cols = 3:12, .fns = as.numeric)) %>% # convert to numeric
  mutate(nu_ns = mean_ns * (1 - mean_ns) / sd_ns^2,
         alpha_ns = mean_ns * nu_ns,
         beta_ns = (1-mean_ns)*nu_ns,
         nu_s = mean_s * (1 - mean_s) / sd_s^2,
         alpha_s = mean_s * nu_s,
         beta_s = (1-mean_s)*nu_s) %>% 
  # set infinite vales to missing
  mutate(across(where(is.numeric), ~ifelse(is.nan(.x), NA_real_, .x))) 
kite_parms %>% 
  select(env_state, age_class, alpha_ns, beta_ns, alpha_s, beta_s) %>% 
  mutate(across(where(is.numeric), ~format(.x, digits = 2))) %>% 
  bind_rows(tibble(env_state = c("","Environmental State"),
                   age_class = c("", "Age Class"),
                   alpha_ns = c("Nest Success", "$\\alpha$"),
                   beta_ns = c("", "$\\beta$"),
                   alpha_s = c("Survival", "$\\alpha$"),
                   beta_s = c("", "$\\beta$")), .) %>% 
  mutate(across(.cols = 3:6, ~case_when(grepl("NA",.x) ~ " ", 
                                        TRUE ~ .x))) %>% 
#  filter(age_class != "Fledgling") %>% 
  hux() %>% 
  set_escape_contents(row = 2, col = 3:6, value = FALSE) %>% 
  theme_article(header_col = FALSE) %>% 
  set_na_string(value = " ") %>% 
  set_position("left") %>% 
  set_caption("Values of $\\alpha$ and $\\beta$ for each age class and 
              environmental state. Fledglings have zero probability of 
              nest success in all environmental states.")
  
```

One of the benefits of articulating the beliefs this way is that the reader can
immediately see how much confidence the analyst has placed in each of the parameter estimates.
The sum of $\alpha$ and $\beta$ is effective sample size. So we have less confidence in the 
estimates of nest success in lag and drought years than high water years, as stated
in Beissinger [-@beissinger1995snailkite]. However, within each environmental 
state we have more confidence in the survival
estimates of fledgling than subadult or adult stages, and less confidence in
estimates from high water years than lag years. Both of these are inconsistent with
the prior beliefs stated by Beissinger[-@beissinger1995snailkite], but are a 
consequence of fixing the standard deviation of the estimates. 
Fixing the standard deviation gives a consistent belief about
precision if the parameter is normally distributed, because the mean and standard deviation
are independent. This is not true of a Beta distribution. 

```{r}
pdf_seq <- seq(-0.2,1.2,0.001)
plot_colors <- RColorBrewer::brewer.pal(3, "RdYlBu")
kite_parms2 <- kite_parms %>% 
  mutate(env_state = rep(env_state[c(1,4,7)], each = 3),
         n_ns = alpha_ns + beta_ns,
         n_s = alpha_s + beta_s,
         lwr_ns = pbeta(min_ns, alpha_ns, beta_ns),
         upr_ns = pbeta(max_ns, alpha_ns, beta_ns),
         lwr_s = pbeta(min_s, alpha_s, beta_s),
         upr_s = pbeta(max_s, alpha_s, beta_s),
         plot_ns = map2(alpha_ns, beta_ns, ~tibble(x_ns = pdf_seq,
                                                   d_ns = dbeta(x_ns, .x, .y))),
         plot_ns_norm = map2(mean_ns, sd_ns, ~tibble(x_nsn = pdf_seq,
                                                     d_ns_norm = dnorm(x_nsn, .x, .y),
                                                     d_ns_tnorm = dnorm(x_nsn, .x, .y)/
                                                       pnorm(0, .x, .y, lower.tail = FALSE))),
         plot_s = map2(alpha_s, beta_s, ~tibble(x_s = pdf_seq,
                                                d_s = dbeta(x_s, .x, .y))),
         plot_s_norm = map2(mean_s, sd_s, ~tibble(x_sn = pdf_seq,
                                                  d_s_norm = dnorm(x_sn, .x, .y),
                                                  d_s_tnorm = dnorm(x_sn, .x, .y)/
                                                    pnorm(1, .x, .y))),
  ) %>% 
  unnest(cols = c("plot_ns", "plot_s", "plot_ns_norm","plot_s_norm"), 
         names_repair = "unique") %>% 
  mutate(env_state = factor(env_state, levels = c("Drought year","Lag year", "High year")),
         age_class = factor(age_class, levels = c("Fledgling", "Subadult", "Adult")))

```

```{r survival-fig, warning=FALSE, fig.cap = "Probability distributions of annual survival by stage and environmental state. Beta distributions are red, truncated normal distributions are blue. Areas of overlap appear gray."}
ggplot(data = kite_parms2) + 
  geom_area(mapping = aes(x = x_s, y = d_s), alpha = 0.75, fill = plot_colors[1], color=plot_colors[1], size = 1) + 
  geom_area(mapping = aes(x = x_sn, y = d_s_tnorm), alpha = 0.75, fill = plot_colors[3], color = plot_colors[3], size = 1) +
  facet_grid(env_state~age_class, scales = "free_x") +
  scale_x_continuous(limits = c(0,1)) + 
  labs(x = "Annual survival", y = "Density")+
  theme_classic()
```

The differences in confidence are not easy to see in the density plots. The
normal distribution is an excellent approximation to the beta distribution
except for subadult and adult stages in high water years. Then the
normal distribution has ~5% probability of values above 1. This was accounted
for by rejecting inadmissible values and drawing another (Beissinger, Personal communication).


## Reproductive rates

Beissinger [-@beissinger1995snailkite] only provides standard deviations for nest success, the
probability of a nest fledging one or more chicks (Table 1). The reproductive rates 
of the matrices are the product of nest success, the number of young fledged per 
successful nest, the proportion of the population nesting, and the number of
nest attempts per year. Other than nest success all of these are treated as
fixed values. A value for the number of fledglings per successful nest was
not provided, but Snyder et al. [-@snyder1989snailkite] had an average of ~ 2 per nest across
all sites and years. Although this value could be used as is, the authors' Table 5
provides actual counts of young per successful nest, 55, 110, and 47 nests having
1, 2 or 3 young respectively. This knowledge is perfectly represented as a 
multinomial variable with a Dirichlet prior. Although the mean success does not
vary across environmental states, the number of nests sampled does, and this
should be reflected in the degree of confidence placed in each distribution. This
representation also includes demographic stochasticity, so we will use a fixed
value of 2 in what follows. 

Beissinger [-@beissinger1995snailkite] used normal distributions for the uncertainty
in nest success, but Table 3 in Snyder et al. [-@snyder1989snailkite] provides
values that can be converted into Beta priors for a binomial variable. 

```{r ns-figure-data}
repro_parms <- tribble(
  # prior_young is vector of nests with 1, 2, or 3 young. 
  # alpha_ns and beta_ns are the number of successful and unsuccessful nests
  ~env_state, ~age_class, ~prior_young, ~alpha_ns, ~beta_ns,
  "Drought year", "Subadult", c(1, 3, 1), 3, 91, 
  "Drought year", "Adult", c(1, 3, 1), 3, 91,
  "Lag year", "Subadult", c(6, 25, 8), 11, 58,
  "Lag year", "Adult", c(6, 25, 8), 11, 58,
  "High year", "Subadult", c(48, 82, 38), 100, 236,
  "High year", "Adult", c(48, 82, 38), 100, 236
)

repro_parms2 <- repro_parms %>% 
  mutate(n_ns = alpha_ns + beta_ns,
         plot_ns = map2(alpha_ns, beta_ns, ~tibble(x = pdf_seq,
                                                   d_ns = dbeta(x, .x, .y)))) %>% 
  unnest(plot_ns) %>% 
  mutate(env_state = factor(env_state, levels = c("Drought year","Lag year", "High year")),
         age_class = factor(age_class, levels = c("Fledgling", "Subadult", "Adult")))

```

```{r ns-figure, warning = FALSE, fig.cap="Prior beliefs about nest success using Beta distributions parameterized directly from Synder et al. (yellow, 1989), Beta distributions parameterized from Beissinger (red, 1995), and normal distributions from Beissinger (blue, 1995)." }

ggplot(data = filter(repro_parms2, age_class != "Fledgling"),
       mapping = aes(x = x, y = d_ns)) + 
  geom_area(fill = plot_colors[2], alpha = 0.75, color=plot_colors[2], size = 1) + #Synder etal 1989
  geom_area(data = filter(kite_parms2, age_class != "Fledgling"), #Beissinger 1995
       mapping = aes(x = x_ns, y = d_ns), fill = plot_colors[1], alpha = 0.75, color=plot_colors[1], size = 1) + 
  geom_area(data = filter(kite_parms2, age_class != "Fledgling"), #Beissinger 1995
       mapping = aes(x = x_ns, y = d_ns_tnorm), fill = plot_colors[3], alpha = 0.75, color=plot_colors[3], size = 1) +   
  facet_grid(env_state~age_class, scales = "free_y") + 
  scale_x_continuous(limits = c(0,0.6)) +
#  scale_y_continuous(limits = c(0,30)) +
  labs(x = "Nest success", y = "Density") +
  rlt_theme
```

The confidence implied by the choices of the mean and variance in Beissinger [-@beissinger1995snailkite] 
is much less than the confidence implied by the sample sizes given in 
Snyder et al. [-@snyder1989snailkite]. It is possible that this is a deliberate
choice because uncertainty in the other components of reproductive success was
assumed to be zero.

Another advantage of using simple count priors for binomial and multinomial
variables is that sampling from those variables automatically obeys restrictions
on the numeric values. Beissinger [-@beissinger1995snailkite] sampled from
normal distributions for survival and nest success. Many of these samples
would have fallen outside the range [0,1] for a probability. Values outside this
range were rejected and resampled (Beissinger, personal communication), giving 
"truncated" normal distributions. For nest success, this creates noticeable
differences only in lag and drought years. In drought years particularly, the 
truncated normal distribution produces nest success values much higher than
either of the parameterizations of the Beta distribution. The mean of the
truncated normal is `r integrate(function(x)x*dnorm(x, mean = 0.03, sd = 0.1)/pnorm(0, mean = 0.03, sd = 0.1, lower.tail = FALSE), lower = 0, upper  = 1)$value`, 3 times larger than the intended
mean.

# Building the matrices

Now that we have distributions for the matrix entries, we can create a function
that returns a matrix with values sampled from the relevant distributions.

```{r}
# fix up parameter dataframes
surv_parms <- kite_parms %>% 
  mutate(env_state = rep(env_state[c(1,4,7)], each = 3)) 
repro_parms <- repro_parms %>% 
  left_join(select(surv_parms, env_state, age_class, prop_nesting, attempts))
snailkite_A <- function(e_state = "High year", 
                        s_parms = surv_parms,
                        r_parms = repro_parms){
  # assumes young per successful nest fixed at 2
  A <- matrix(0, nrow = 3, ncol = 3)
  surv_parms <- filter(s_parms, env_state == e_state)
  repro_parms <- filter(r_parms, env_state == e_state)
  surv <- rbeta(3, surv_parms$alpha_s, surv_parms$beta_s)
  A[2,1] <- surv[2]
  A[3,2] <- A[3,3] <- surv[3]
  ns <- rbeta(2, repro_parms$alpha_ns, repro_parms$beta_ns)
  A[1,2:3] <- repro_parms$prop_nesting * repro_parms$attempts * ns * 2 * surv[1]
  A
}
snailkite_A()
snailkite_A("Drought year")
snailkite_A("Lag year")
```

We can then draw many samples of each matrix to obtain credible intervals
on derived quantities like the asymptotic growth rate $\lambda$. 

```{r}
results <- tibble(
  e_state = factor(rep(c("Drought year", "Lag year", "High year"), each = 100),
                   levels = c("Drought year", "Lag year", "High year")),
  A = map(e_state, snailkite_A),
  lambda = map_dbl(A, lambda)
)
ggplot(data = results,
       mapping = aes(x = lambda)) + 
  geom_histogram(binwidth = 0.05) + 
  facet_wrap(~e_state) +
  rlt_theme
```

We can reproduce extinction risk plots as in Beissinger [-@beissinger1995snailkite],
if that is necessary. 

# Literature cited
